Designing Types for R, Empirically

### 2020-05-16 paper/main.tex
merge

\newcommand{\NUMFUNCTIONS}{20,214\xspace}  %%% TODO auto-gen
\newcommand{\NUMPACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\PACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\genthat}{genthat\xspace}  %%% TODO auto-gen
\newcommand{\YEARS}{20\xspace} %%TODO fix
\newcommand{\PERCFAILEDASSERTIONS}{2.19\%\xspace}
\newcommand{\PERCFAILEDASSERTIONSWUNDEF}{2.23\%\xspace}
\newcommand{\PERCASSERTIONSUNDEF}{0.04\%\xspace}
\newcommand{\NUMPKGSEVAL}{7485\xspace} % autogenerate
\newcommand{\NUMASSERTIONS}{62,171,573\xspace} %autogenerate
\newcommand{\PROPFUNSFAILEDCHECK}{21.21\%\xspace}
\newcommand{\PROPFUNSFAILEDCHECKNOSTHREE}{15.48\%\xspace} 
\newcommand{\PROPARGSFAILINGASSERTS}{3.53\%\xspace} % autogen....
\newcommand{\PERCSUCCARG}{96.46\%\xspace}
\newcommand{\PERCSUCCFUNNOSTHREE}{84.52\%\xspace}
\newcommand{\PERCSUCCFUNS}{78.79\%\xspace}
\newcommand{\PERCCALLSBADFUNSINTESTS}{5.38\%\xspace}
\newcommand{\TOTNUMSIGS}{21,730\xspace}
\newcommand{\PROPSTRUCTSDYN}{44.46\%\xspace}

@@

To sum up, our paper makes the following main contributions:
\begin{itemize}
\item We implemented tooling to automatically extract type signatures from R
  functions and to instrument R functions with checks based on their declared
  types. The tooling is robust and scalable to the entire R language.
\item We carried out a large-scale analysis of corpus of \PACKAGES widely
  used and actively maintained libraries to extract function type signatures
  validated the robustness of the inferred type signatures against XXX
  programs that use those functions.
\item We report on the appropriateness and usefulness of a simple type
  language for the R programming language.
\end{itemize}

@@

\noindent
Assuming that type sequences can be reordered freely, we rewrite types to
minimize their size by removing redundant types, types that are subsumed by
subtyping, and \null types.

@@

We ran \typetracer on the \CorpusLoadable packages we discussed in this section, which yields types for \NUMFUNCTIONS functions.
To build up a picture of what types for R functions might look like, consult Table~\ref{tbl:corpus-type-breakdown}, which lists the 10 most popular argument signatures.
The first row of the table reads: 13255 arguments have type \dbl, represnting 11.06\% of type signatures, and \typetracer observed 41M values of type \dbl flowing into or out of functions while running on the core package corpus, which represents only 1.59\% of dynamically observed values.
Of interest are rows 2 and 4, where we see that collectively \lgl and \chr make up only 16.11\% of argument types, while making up nearly 60\% of dynamically observed values.

@@

\begin{tabular}{lrrrr}
\toprule
Type & Args with Type & \% of Static Types & Dynamic Observations & \% of Dynamic Types\\
\midrule
\dbl & 13,255 & 11.06 & 41,560,797 & 1.59\\
\lgl & 10,142 & 8.46 & 712,895,959 & 27.21\\
\null & 9,175 & 7.66 & 122,202,183 & 4.66\\
\chr & 9,171 & 7.65 & 794,058,588 & 30.31\\
\VEC\dbl & 7,811 & 6.52 & 29,272,232 & 1.12\\
\addlinespace
\VARGS & 7,759 & 6.48 & 76,369,333 & 2.92\\
\any & 6,243 & 5.21 & 255,539,497 & 9.75\\
\VEC\chr & 4,576 & 3.82 & 88,233,523 & 3.37\\
\CLASS{matrix} & 4,522 & 3.77 & 32,452,555 & 1.24\\
\CLASS{data.frame} & 2,781 & 2.32 & 7,749,094 & 0.30\\
\bottomrule
\end{tabular}

@@

>>>>

\newcommand{\NUMFUNCTIONS}{20,214\xspace}  %%% TODO auto-gen
\newcommand{\NUMPACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\PACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\genthat}{genthat\xspace}  %%% TODO auto-gen
\newcommand{\YEARS}{20\xspace} %%TODO fix
\newcommand{\PERCFAILEDASSERTIONS}{2.19\%\xspace}
\newcommand{\PERCFAILEDASSERTIONSWUNDEF}{2.23\%\xspace}
\newcommand{\PERCASSERTIONSUNDEF}{0.04\%\xspace}
\newcommand{\NUMPKGSEVAL}{7485\xspace} % autogenerate
\newcommand{\NUMASSERTIONS}{62,171,573\xspace} %autogenerate
\newcommand{\PROPFUNSFAILEDCHECK}{21.21\%\xspace}
\newcommand{\PROPFUNSFAILEDCHECKNOSTHREE}{15.48\%\xspace} 
\newcommand{\PROPARGSFAILINGASSERTS}{3.53\%\xspace} % autogen....
\newcommand{\PERCSUCCARG}{96.46\%\xspace}
\newcommand{\PERCSUCCFUNNOSTHREE}{84.52\%\xspace}
\newcommand{\PERCSUCCFUNS}{78.79\%\xspace}
\newcommand{\PERCCALLSBADFUNSINTESTS}{5.38\%\xspace}
\newcommand{\TOTNUMSIGS}{21,730\xspace}
\newcommand{\PROPSTRUCTSDYN}{44.46\%\xspace}
\newcommand{\NUMFILESRUN}{145,373\xspace}

@@

To sum up, our paper makes the following main contributions:
\begin{itemize}
\item We implemented tooling to automatically extract type signatures from R
  functions and to instrument R functions with checks based on their declared
  types. The tooling is robust and scalable to the entire R language.
\item We carried out a large-scale analysis of corpus of \PACKAGES widely
  used and actively maintained libraries to extract function type signatures and
  validated the robustness of the inferred type signatures against \NUMFILESRUN
  programs that use those functions.
\item We report on the appropriateness and usefulness of a simple type
  language for the R programming language.
\end{itemize}

@@

\noindent
Assuming that type sequences can be reordered freely, we rewrite types to
minimize their size by removing redundant types, types that are subsumed by
subtyping, and \null types. Lists with more than 5 types in their union become \LIST{any}.
We also only generate types for the first 20 arguments of a function.

@@

We ran \typetracer on the \CorpusLoadable packages we discussed in this section, which yields types for \NUMFUNCTIONS functions.
To build up a picture of what types for R functions might look like, consult Table~\ref{tbl:corpus-type-breakdown}, which lists the 10 most popular argument signatures.
The first row of the table reads: 13255 arguments have type \dbl, represnting 11.06\% of argument type signatures, and \typetracer observed 41M values of type \dbl flowing into or out of functions while running on the core package corpus, which represents only 1.59\% of dynamically observed values.
Of interest are rows 2 and 4, where we see that collectively \lgl and \chr make up only 16.11\% of argument types, while making up nearly 60\% of dynamically observed values.

@@

\begin{tabular}{lrrrr}
\toprule
Type & Args with Type & \% Args & Dynamic Observations & \% Dynamic Types\\
\midrule
\dbl & 13,255 & 11.06 & 41,560,797 & 1.59\\
\lgl & 10,142 & 8.46 & 712,895,959 & 27.21\\
\null & 9,175 & 7.66 & 122,202,183 & 4.66\\
\chr & 9,171 & 7.65 & 794,058,588 & 30.31\\
\VEC\dbl & 7,811 & 6.52 & 29,272,232 & 1.12\\
\addlinespace
\VARGS & 7,759 & 6.48 & 76,369,333 & 2.92\\
\any & 6,243 & 5.21 & 255,539,497 & 9.75\\
\VEC\chr & 4,576 & 3.82 & 88,233,523 & 3.37\\
\CLASS{matrix} & 4,522 & 3.77 & 32,452,555 & 1.24\\
\CLASS{data.frame} & 2,781 & 2.32 & 7,749,094 & 0.30\\
\bottomrule
\end{tabular}

@@

\section{Future Work}

We see two major avenues for future work.

\subsection{Types for R Objects}

\subsection{Types for Data Frames}
